### 2016-04-19 Manuscript/SW_early_life_telomeres.Rmd
DSR edits

We used linear mixed models with model averaging to explore how spatial variation in early-life environmental and social conditions influenced telomere length within cohorts. Model averaging was carried out using the MuMIn package (version 1.10.5) in R [@Barton2012]. We first created a full model containing the following explanatory variables: age (in months), tarsus length, sex, territory quality and the number of helping and non-helping subordinate birds present in the natal territory. We also included interaction terms between age and all the other variables as telomere dynamics are expected to vary within the first year of life [@Heidinger2012]. As random effects we included cohort, territory ID and qPCR plate ID. Model selection was then performed and a top model set defined, containing all models with AICc $\leq$ 6 compared to the best supported model [@Burnham2011]. We report model-averaged coefficients, confidence intervals and 'relative importance', which reflects the relative weights of each predictor variable across the top model set. For individuals with longitudinal data (*n* = `r nrow(Loss)`) we repeated the above analyses of telomere dynamics, replacing telomere length with telomere loss as the response variable, and excluding the plate ID random effect (as each measurement of telomere loss was based on two or more measurements, and so run on multiple plates). For both sets of models, we first ran the model averaging procedure including all interaction terms, then re-ran the pipeline excluding non-significant interactions.

@@

$\Delta$RTL did not differ significantly among cohorts of nestlings (F = `r round2(anova(lm(Loss~factor(FieldPeriodID),data = subset(Loss,Agemonths == 1)))[1,4])`; P `r round3(anova(lm(Loss~factor(FieldPeriodID),data = subset(Loss,Agemonths == 1)))[1,5])`), although power for this analysis was limited as longitudinal sample sizes within cohorts were low. Model selection revealed that the top model explaining $\Delta$RTL within cohorts of nestlings contained sex and tarsus length (Table S2). This model was a better fit than the null model ($\Delta$AICc = `r round2(AICc(lmer(Loss~1+(1|FieldPeriodID)+(1|TerritoryID),data=subset(Loss,Agemonths == 1),REML = FALSE))-AICc(lmer(Loss~Sex+Tarsus+(1|FieldPeriodID)+(1|TerritoryID),data=subset(Loss,Agemonths == 1),REML = FALSE)))`), although confidence limits for both the sex and tarsus effects overlapped zero (Fig. 5A). $\Delta$RTL was negatively related to tarsus length, and higher in females, indicating greater telomere shortening in smaller nestlings, and in males (Fig. 5B).

@@

Early-life RTL was not related to survival to adulthood (estimate = `r round2(summary(glmer(SurvivedNext~RTL+Agemonths+(1|TerritoryID)+(1|FieldPeriodID),data=juv13,family = 'binomial'))$coefficients[2,1])`; P = `r round2(summary(glmer(SurvivedNext~RTL+Agemonths+(1|TerritoryID)+(1|FieldPeriodID),data=juv13,family = 'binomial'))$coefficients[2,4])`), with no interaction effect between RTL and juvenile age (estimate = `r round2(summary(glmer(SurvivedNext~RTL*Agemonths+(1|TerritoryID)+(1|FieldPeriodID),data=juv13,family = 'binomial'))$coefficients[4,1])`; P = `r round2(summary(glmer(SurvivedNext~RTL*Agemonths+(1|TerritoryID)+(1|FieldPeriodID),data=juv13,family = 'binomial'))$coefficients[4,4])`). To test whether cohort-level effects on RTL influenced survival we calculated cohort-level RTL (i.e. the median RTL for each cohort) and added this term to a GLMM, and found that cohort-level RTL was not related to survival to adulthood (estimate = `r round2(summary(glmer(SurvivedNext~cohortTL+RTL+Agemonths+(1|TerritoryID)+(1|FieldPeriodID),data=juv13,family = 'binomial'))$coefficients[2,1])`; P = `r round2(summary(glmer(SurvivedNext~cohortTL+RTL+Agemonths+(1|TerritoryID)+(1|FieldPeriodID),data=juv13,family = 'binomial'))$coefficients[2,4])`). Using the longitudinal data, we found that $\Delta$RTL had no effect on survival to the year after the second sampling event (estimate = `r round2(summary(glm(SurvivedNext2~Loss+Agemonths,data = subset(Loss,FieldPeriodID<2013),family = 'binomial'))$coefficients[2,1])`; P = `r round2(summary(glm(SurvivedNext2~Loss+Agemonths,data=subset(Loss,LayYear<2013),family = 'binomial'))$coefficients[2,4])`).

@@

###Acknowledgements
We thank Nature Seychelles for facilitating the long-term Seychelles warbler project. The Seychelles Bureau of Standards and Department of Environment gave permission for sampling and fieldwork. Emma Barrett laid the foundations for this study, generating the original telomere qPCR protocol. We thank everyone who has helped in the field, and the Seychelles warbler research group for discussions. This work was funded by two Natural Environment Research Council (NERC) grants to DSR (NE/F02083X/1 and NE/K005502/1). LGS was also funded by an EGI Fellowship from the University of Oxford, and HLD by a NERC fellowship.

>>>>

We used linear mixed models with model averaging to explore how spatial variation in early-life environmental and social conditions influenced telomere length within cohorts. Model averaging was carried out using the MuMIn package (version 1.10.5) in R [@Barton2012]. We first created a full model containing the following explanatory variables: age (in months), tarsus length, sex, territory quality and the number of helping and non-helping subordinate birds present in the natal territory. We also included interaction terms between age and all the other variables as telomere dynamics are expected to vary within the first year of life [@Heidinger2012]. As random effects we included cohort, territory ID and qPCR plate ID. Model selection was then performed and a top model set defined, containing all models with AICc $\leq$ 6 compared to the best supported model [@Burnham2011]. We report model-averaged coefficients, confidence intervals and 'relative importance', which reflects the relative weights of each predictor variable across the top model set. For individuals with longitudinal data (*n* = `r nrow(Loss)`) we repeated the above analyses of telomere dynamics, replacing telomere length with telomere loss as the response variable, and excluding the plate ID random effect (as each measurement of telomere loss was based on two or more measurements, and so run on multiple plates). For both sets of models, we first ran the model averaging procedure including all interaction terms, then re-ran the pipeline excluding non-significant interactions in order to obtain more accurate model averaged parameter estimates and confidence intervals.

@@

$\Delta$RTL did not differ significantly among cohorts (F = `r round2(anova(lm(Loss~factor(FieldPeriodID),data = Loss))[1,4])`; P `r round3(anova(lm(Loss~factor(FieldPeriodID),data = Loss))[1,5])`), although power for this analysis was limited as longitudinal sample sizes within cohorts were low. Model selection revealed that the top model explaining $\Delta$RTL within cohorts contained only age (Table S2). This model was a better fit than the null model ($\Delta$AICc = `r round2(AICc(lmer(Loss~1+(1|FieldPeriodID)+(1|TerritoryID),data=Loss,REML = FALSE))-AICc(lmer(Loss~Agemonths+(1|FieldPeriodID)+(1|TerritoryID),data=Loss,REML = FALSE)))`), and only the age effect was significant (Fig. 4C).

@@

Early-life RTL was not related to survival to adulthood (estimate = `r round2(summary(glmer(SurvivedNext~RTL+(1|FieldPeriodID),data=subset(juv13,Agemonths ==1),family = 'binomial'))$coefficients[2,1])`; P = `r round2(summary(glmer(SurvivedNext~RTL+Agemonths+(1|TerritoryID)+(1|FieldPeriodID),data=juv13,family = 'binomial'))$coefficients[2,4])`), with no interaction effect between RTL and juvenile age (estimate = `r round2(summary(glmer(SurvivedNext~RTL*Agemonths+(1|TerritoryID)+(1|FieldPeriodID),data=juv13,family = 'binomial'))$coefficients[4,1])`; P = `r round2(summary(glmer(SurvivedNext~RTL*Agemonths+(1|TerritoryID)+(1|FieldPeriodID),data=juv13,family = 'binomial'))$coefficients[4,4])`). To test whether cohort-level effects on RTL influenced survival we calculated cohort-level RTL (i.e. the median RTL for each cohort) and added this term to a GLMM, and found that cohort-level RTL was not related to survival to adulthood (estimate = `r round2(summary(glmer(SurvivedNext~cohortTL+RTL+Agemonths+(1|TerritoryID)+(1|FieldPeriodID),data=juv13,family = 'binomial'))$coefficients[2,1])`; P = `r round2(summary(glmer(SurvivedNext~cohortTL+RTL+Agemonths+(1|TerritoryID)+(1|FieldPeriodID),data=juv13,family = 'binomial'))$coefficients[2,4])`). Using the longitudinal data, we found that $\Delta$RTL had no effect on survival to the year after the second sampling event (estimate = `r round2(summary(glm(SurvivedNext2~Loss+Agemonths,data = subset(Loss,FieldPeriodID<2013),family = 'binomial'))$coefficients[2,1])`; P = `r round2(summary(glm(SurvivedNext2~Loss+Agemonths,data=subset(Loss,LayYear<2013),family = 'binomial'))$coefficients[2,4])`).

@@

###Acknowledgements
We thank Nature Seychelles for facilitating the long-term Seychelles warbler project. The Seychelles Bureau of Standards and Department of Environment gave permission for sampling and fieldwork. Emma Barrett laid the foundations for this study, generating the original telomere qPCR protocol. We thank everyone who has helped in the field, and the Seychelles warbler research group for discussions. This work was funded by two Natural Environment Research Council (NERC) grants to DSR (NE/F02083X/1 and NE/K005502/1). LGS was also funded by fellowships from the University of Oxford and the BBSRC, and HLD by a NERC fellowship.

### 2016-04-19 Manuscript/SW_early_life_telomeres_SOM.Rmd
DSR edits

```{r Table S2}
temp <- subset(Loss,Agemonths == 1)

#Run the model
lm1 <- lmer(Loss ~ (Tarsus + Helper + NonHelper + Sex + BodyMass  + TQ) + (1|FieldPeriodID),
            data = temp,
            REML = FALSE)
lm1 <- standardize(lm1)

@@

nams[nams == 'z.RTL'] <- 'RTL'
nams[nams == 'z.NonHelper'] <- 'Non helpers'
nams[nams == 'c.Sex'] <- 'Sex'
nams[nams == 'z.TQ'] <- 'TQ'
nams[nams == 'c.Helper'] <- 'Helpers'
nams[nams == 'z.BodyMass'] <- 'Body mass'
nams[nams == 'z.Agemonths'] <- 'Age'
nams[nams == 'z.Tarsus'] <- 'Tarsus'
nams[nams == 'c.Season'] <- 'Season'
nams[nams == 'c.Season:z.Agemonths'] <- 'Age x season'
nams[nams == 'z.Agemonths:z.NonHelper'] <- 'Age x non helpers'
nams[nams == 'c.Sex:z.Agemonths'] <- 'Age x sex'
nams[nams == 'z.Agemonths:z.TQ'] <- 'Age x TQ'
nams[nams == 'c.Helper:z.Agemonths'] <- 'Age x helpers'
nams[nams == 'z.Agemonths:z.BodyMass'] <- 'Age x mass'
nams[nams == 'z.Agemonths:z.Tarsus'] <- 'Age x tarsus'

@@

tableS2 <- data.frame(Model = apply(t1,1,getterms,c(2:7)),t1[,c('df','AICc','delta','weight')],stringsAsFactors = FALSE)
colnames(tableS2) <- c('Model','df','AICc', 'Delta AICc','Weight')
rownames(tableS2) <- NULL
tableS2$Model[tableS2$Model == ''] <- 'Null model'
tableS2[,3:5] <- lapply(tableS2[,3:5],as.numeric)
kable(tableS2,digits=2)

>>>>

```{r Table S2}
temp <- Loss

#Run the model
lm1 <- lmer(Loss ~ Agemonths*(Tarsus + Helper + NonHelper + Sex + BodyMass  + TQ) + (1|FieldPeriodID),
            data = temp,
            REML = FALSE)
lm1 <- standardize(lm1)

@@

@@

### 2016-04-19 Manuscript/SW_early_life_telomeres_tables_and_figures.Rmd
DSR edits

Fig4A <- ggplot(temp,
       aes(x = RTL,
           y = RTL1,
           col=`Age (months)`)) +
  ylab('Adult RTL') + 
  xlab('Juvenile RTL') +
  theme_classic() +
  theme(axis.text.x = element_text(size = 12), 
        axis.text.y = element_text(size = 12), 
        axis.title.y = element_text(size = 14,vjust=0.8),
        axis.title.x = element_text(size = 14),
        legend.position = 'top') +
  geom_point()+
  scale_colour_manual(values = c('grey','darkorange','darkgreen'))+
  scale_fill_manual(values = c('grey','darkorange','darkgreen'))
  annotate('text',x = 0,y = 2.1,label = 'A')

@@

pdf('Figures/Figure 4.pdf',
    width = 4,
    height = 8,
    paper = 'a4')
multiplot(Fig4A,Fig4B,layout=matrix(c(1,2),nrow=2,byrow=T))
dev.off()

```{r Figure 5}

temp <- Loss
#Run the model
lm2 <- lmer(Loss ~ TimeDiff+Agemonths*Helper+Tarsus + NonHelper + Sex + BodyMass  + TQ + (1|FieldPeriodID),
            data = temp,
            REML = FALSE)
lm2 <- standardize(lm2)

@@

Fig5A <- forestplot(table2)+
  annotate('text',x = 0.75,y = 0.6,label = 'A')

temp <- Loss
temp[,'Helpers'] <- ifelse(temp$Helper>0,'Present', 'Absent')
temp[,'Non helpers'] <- ifelse(temp$NonHelper>0,'Present', 'Absent')
temp[,'Age (months)'] <- factor(temp$Agemonths)

Fig5B <- ggplot(temp,
       aes(x = `Age (months)`,
           y = Loss,
           fill=Helpers)) +
  theme_classic() +
  ylab(expression(Delta*'RTL'))+
  theme(axis.text.x = element_text(size = 12), 
        axis.text.y = element_text(size = 12), 
        axis.title.y = element_text(size = 14,vjust=0.8),
        axis.title.x = element_text(size = 14),
        legend.position = 'top') +
  geom_jitter(aes(colour = Helpers),position=position_jitterdodge(dodge.width = 0.8))+
  geom_boxplot(outlier.colour = NA,notch=F,alpha=0.2)+
  scale_colour_manual(values = c('grey','darkorange'))+
  scale_fill_manual(values = c('grey','darkorange'))+
  geom_abline(slope = 0,intercept = 0,lty = 2)+
  annotate('text',x = 0.55,y = 1.5,label = 'B')

Fig5C <- ggplot(temp,
       aes(col = `Age (months)`,
           fill =  `Age (months)`,
           y = Loss,
           x=TQ)) +
    ylab(expression(Delta*'RTL'))+
  theme_classic() +
  theme(axis.text.x = element_text(size = 12), 
        axis.text.y = element_text(size = 12), 
        axis.title.y = element_text(size = 14,vjust=0.8),
        axis.title.x = element_text(size = 14),
        legend.position = 'top') +
  geom_point()+
  stat_smooth(method = 'lm')+
  scale_colour_manual(values = c('grey','darkorange','darkgreen'))+
  scale_fill_manual(values = c('grey','darkorange','darkgreen'))+
  annotate('text',x = 3,y = 1.3,label = 'C')

pdf('Figures/Figure 5.pdf',
    width = 8,
    height = 8,
    paper = 'a4')
multiplot(Fig5A,Fig5B,Fig5C,layout = matrix(c(1,1,2,3),nrow=2,byrow = T))
dev.off()

@@

>>>>

Fig4A <- ggplot(temp,
       aes(x = RTL,
           y = RTL1,
           col=`Age (months)`)) +
  ylab('Adult RTL') + 
  xlab('Juvenile RTL') +
  theme_classic() +
  theme(axis.text.x = element_text(size = 12), 
        axis.text.y = element_text(size = 12), 
        axis.title.y = element_text(size = 14,vjust=0.8),
        axis.title.x = element_text(size = 14),
        legend.position = c(0.9,0.9),
        legend.background = element_rect(fill = grey(0.9))) +
  geom_point()+
  scale_colour_manual(values = c('grey','darkorange','darkgreen'))+
  scale_fill_manual(values = c('grey','darkorange','darkgreen'))+
  annotate('text',x = 0,y = 2.1,label = 'A')

@@

temp <- Loss
#Run the model
lm2 <- lmer(Loss ~ Agemonths+Helper+Tarsus + NonHelper + Sex + BodyMass  + TQ + (1|FieldPeriodID),
            data = temp,
            REML = FALSE)
lm2 <- standardize(lm2)

@@

Fig4C <- forestplot(table2)+
  annotate('text',x = 0.75,y = 0.6,label = 'C')

pdf('Figures/Figure 4.pdf',
    width = 8,
    height = 8,
    paper = 'a4')
multiplot(Fig4A,Fig4B,Fig4C,layout=matrix(c(1,2,3,3),nrow=2,byrow=T))
dev.off()

@@

```{r Figure 5}
temp <- subset(juv13,Agemonths != 1)
temp$LayYear <- factor(temp$LayYear)
temp[,'Survived to following year?'] <- ifelse(temp$SurvivedNext == 1,'Y','N')

Fig4B <- ggplot(temp,
       aes(x = LayYear,
           y = RTL,
           fill = `Survived to following year?`)) +
  xlab('Cohort') +
  theme_classic() +
  theme(axis.text.x = element_text(size = 12), 
        axis.text.y = element_text(size = 12), 
        axis.title.y = element_text(size = 14,vjust=0.8),
        axis.title.x = element_text(size = 14),
        legend.position = 'top') +
  geom_boxplot(outlier.shape = NA,alpha = 0.2)+
  geom_jitter(aes(col = `Survived to following year?`),position=position_jitterdodge(jitter.width = 0.2))+
  annotate('text',x = 0.5,y = 1.5,label = 'B')

Fig4B
